---
title: "Correlated evolution of module selection and hair dichromatism"
subtitle: "Opie-analog phylogenetic tests — frozen, replicable"
author: "Molecular evolution analysis (pigmentation-gene-network)"
format:
  html:
    toc: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

## What this notebook does

Three phylogenetic tests on **frozen inputs** in `./data/` — no network, no
external state. `quarto render` (or run each chunk) reproduces every number and
the figure. This is the companion to `../module_selection/`: that notebook builds
the per-lineage module-balance metric and the Opie ancestral-state /
transition-rate analysis; **this** notebook adds the one Opie test not covered
there — the **correlated-evolution** test (Pagel's *Discrete*, ML analog via
`phytools::fitPagel`) — plus Pagel's lambda for the trait.

The Opie, Atkinson & Shultz (2012, *Commun Integr Biol* 5:458,
doi:10.4161/cib.20821) analysis of primate **mating systems** used three tools:
phylogenetic signal / model choice, ancestral-state reconstruction, and a test
of whether two discrete traits evolve **dependently** (correlated) rather than
independently. We hold no mating-system data, so we replicate their *method* on
our own phenotype + selection results.

**Question here:** does a lineage being under selection in the pigmentation
module (or the sex-hormone module) co-evolve with that lineage being
dichromatic? This is the lineage-level version of the project's set-level
pigment-vs-hormone contrast (which was not significant, permutation p=0.87).

### Data provenance (frozen)

| file | source | contents |
|---|---|---|
| `data/branch_rates.csv` | HPC full-panel aBSREL, commit 4c07317 (`results/perorigin_v1/`) | 78 genes x 288 branches; per-branch omega, corrected p, `selected_flag` |
| `data/gene_modules.csv` | curated panel | gene -> module (pigmentation / hormone) |
| `data/species_coding.csv` | genomic-panel phenotype | 117 species -> dichromatic 0/1 |
| `data/primate_species_tree.nex` | pruned genomic tree | 304-tip tree for the genome-sampled set |
| `data/leakey_pruned_tree.nex` | Leakey2025 phenotype tree | 235-tip primate tree |
| `data/leakey_dichromatism_coding.csv` | Leakey2025 `primate_prelim_complete.csv` (2 cols) | species_binom, hair_dichromatism_any |

The **lambda and transition-rate** tests use the 235-tip Leakey phenotype tree
(not the genome-sampled set, which is ascertainment-biased toward dichromatic
species). The **fitPagel** test needs both traits per tip, so it uses the
genomic panel (species with an aBSREL result).

```{r setup}
library(ape); library(phytools); library(tidyverse)
dd <- "data"
BR  <- read_csv(file.path(dd, "branch_rates.csv"), show_col_types = FALSE)
GM  <- read_csv(file.path(dd, "gene_modules.csv"), show_col_types = FALSE)
SC  <- read_csv(file.path(dd, "species_coding.csv"), show_col_types = FALSE)
gtr <- read.nexus(file.path(dd, "primate_species_tree.nex"))
ltr <- read.nexus(file.path(dd, "leakey_pruned_tree.nex"))
LC  <- read_csv(file.path(dd, "leakey_dichromatism_coding.csv"), show_col_types = FALSE)
```

## 1. Pagel's lambda — phylogenetic signal in dichromatism

Opie reported lambda = 0.996 for mating systems (near-maximal signal). We
estimate lambda for hair dichromatism on the Leakey phenotype tree, treating the
0/1 trait as continuous (a standard proxy for a binary trait), with a
likelihood-ratio test against lambda = 0.

```{r lambda}
ph <- LC %>% mutate(sp = gsub(" ", "_", species_binom)) %>%
  select(sp, dich = hair_dichromatism_any) %>% filter(!is.na(dich))
common <- intersect(ltr$tip.label, ph$sp)
trm <- keep.tip(ltr, common)
x   <- ph %>% filter(sp %in% common) %>% distinct(sp, .keep_all = TRUE)
xv  <- setNames(as.integer(x$dich), x$sp)[trm$tip.label]

ps <- phylosig(trm, xv, method = "lambda", test = TRUE)
lambda_res <- tibble(n_tips = length(common), n_dichromatic = sum(xv),
                     lambda = ps$lambda, logL = ps$logL, logL0 = ps$logL0, p = ps$P)
lambda_res
```

## 2. Transition-rate asymmetry (loss vs gain) — cross-check

The ancestral-state / transition-rate analysis is the primary output of the
sibling `module_selection/` notebook; we recompute the ARD-vs-ER model choice
here as an independent cross-check. **Note:** `ace(model="ARD")$rates` does not
return rates in gain-then-loss order — confirm direction with `fitMk` ->
`as.Qmatrix()` (rows = from-state, cols = to-state).

```{r rates}
fm  <- fitMk(trm, setNames(as.character(xv), names(xv)), model = "ARD")
Qm  <- unclass(as.Qmatrix(fm))
fit_er  <- ace(xv, trm, type = "discrete", model = "ER")
fit_ard <- ace(xv, trm, type = "discrete", model = "ARD")
rate_res <- tibble(
  gain_0to1 = Qm[1, 2], loss_1to0 = Qm[2, 1], loss_over_gain = Qm[2, 1] / Qm[1, 2],
  AIC_ER  = 2 * 1 - 2 * fit_er$loglik,
  AIC_ARD = 2 * 2 - 2 * fit_ard$loglik) %>%
  mutate(dAIC_ARD_minus_ER = AIC_ARD - AIC_ER)
rate_res
```

## 3. fitPagel — does module selection co-evolve with dichromatism?

Two binary traits per tip: (i) dichromatic vs monochromatic; (ii) at least one
gene of the module under aBSREL episodic selection (corrected p < 0.05) on that
tip's terminal branch. `fitPagel` compares an **independent** model (the two
traits evolve separately) against a **dependent** model (each trait's transition
rates depend on the other's state) — the ML analog of Opie's BayesTraits
*Discrete* dependent-vs-independent Bayes-factor test.

```{r fitpagel}
sel_tip <- BR %>% filter(is_tip == TRUE, selected_flag == TRUE) %>%
  distinct(branch, set)
pig_tips <- sel_tip %>% filter(set == "pigmentation") %>% pull(branch)
hor_tips <- sel_tip %>% filter(set == "hormone") %>% pull(branch)

tested_tips <- unique(BR$branch[BR$is_tip == TRUE])
keep <- Reduce(intersect, list(gtr$tip.label, SC$species, tested_tips))
gtrm <- keep.tip(gtr, keep)

dich <- setNames(ifelse(SC$dichromatic[match(keep, SC$species)] == 1, "dich", "mono"), keep)
pigs <- setNames(ifelse(keep %in% pig_tips, "pig_sel", "none"), keep)
hors <- setNames(ifelse(keep %in% hor_tips, "hor_sel", "none"), keep)

set.seed(1)
fp_pig <- fitPagel(gtrm, dich, pigs)
fp_hor <- fitPagel(gtrm, dich, hors)

pagel_res <- tibble(
  module = c("pigmentation", "hormone"),
  n_tips = length(keep), n_dichromatic = sum(dich == "dich"),
  indep_AIC = c(fp_pig$independent.AIC, fp_hor$independent.AIC),
  dep_AIC   = c(fp_pig$dependent.AIC,   fp_hor$dependent.AIC),
  LR = c(fp_pig$lik.ratio, fp_hor$lik.ratio),
  p  = c(fp_pig$P, fp_hor$P)) %>%
  mutate(dAIC_indep_minus_dep = indep_AIC - dep_AIC,
         preferred = ifelse(dep_AIC < indep_AIC, "dependent", "independent"))
pagel_res
```

### Figure

```{r figure, fig.width=7.2, fig.height=4.6}
df <- pagel_res %>%
  select(module, indep_AIC, dep_AIC, p) %>%
  pivot_longer(c(indep_AIC, dep_AIC), names_to = "model", values_to = "AIC") %>%
  mutate(model = factor(ifelse(model == "indep_AIC", "Independent", "Dependent"),
                        levels = c("Independent", "Dependent")),
         module = factor(ifelse(module == "pigmentation", "Pigmentation module",
                                "Sex-hormone module"),
                         levels = c("Pigmentation module", "Sex-hormone module")))
lab <- pagel_res %>%
  transmute(module = factor(ifelse(module == "pigmentation", "Pigmentation module",
                                   "Sex-hormone module"),
                            levels = c("Pigmentation module", "Sex-hormone module")),
            y = pmax(indep_AIC, dep_AIC) + 2.5,
            txt = sprintf("independent preferred\ndAIC=%.1f, LR p=%.2f",
                          dAIC_indep_minus_dep, p))
p <- ggplot(df, aes(module, AIC, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.62) +
  geom_text(aes(label = sprintf("%.1f", AIC)), position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3.1, colour = "grey20") +
  geom_text(data = lab, aes(module, y, label = txt), inherit.aes = FALSE,
            size = 3.0, colour = "grey25", lineheight = 0.95, vjust = 0) +
  scale_fill_manual(values = c(Independent = "#8C8C8C", Dependent = "#3B3B3B"),
                    name = "Model of evolution") +
  coord_cartesian(ylim = c(min(df$AIC) - 8, max(df$AIC) + 9)) +
  labs(title = "Module selection does not co-evolve with dichromatism",
       subtitle = sprintf("fitPagel, %d-tip genomic panel (%d dichromatic); Pagel's lambda=%.2f, p=%.1g for the trait itself",
                          nrow(SC[SC$species %in% keep, ]), sum(dich == "dich"),
                          lambda_res$lambda, lambda_res$p),
       x = NULL, y = "AIC (lower = better fit)") +
  theme_classic(base_size = 11) +
  theme(legend.position = "top", plot.title = element_text(face = "plain", size = 12),
        plot.subtitle = element_text(size = 8.3, colour = "grey35"),
        axis.text.x = element_text(size = 10))
ggsave("fig_fitpagel_correlated_evolution.png", plot = p, width = 7.2, height = 4.6, dpi = 300)
p
```

## 4. Findings

- **Dichromatism carries strong phylogenetic signal** (Pagel's lambda ~ 0.65,
  LR p < 1e-6) — closely related species resemble each other in the trait, as
  expected and consistent with Opie's near-maximal lambda for mating systems.
- **Loss exceeds gain ~9x** (ARD asymmetric model preferred over equal-rates,
  dAIC ~ 20) — dichromatism is evolutionarily labile, lost far more readily than
  gained. This cross-check reproduces the `module_selection/` result.
- **Module selection does NOT co-evolve with dichromatism.** For both the
  pigmentation and sex-hormone modules the **independent** model is AIC-preferred
  (dAIC 3.7 and 5.7; LR p 0.37 and 0.68). There is no lineage-level statistical
  coupling between which module is under selection and whether a lineage is
  dichromatic.

**Interpretation & power caveat.** The negative fitPagel result reinforces the
project's non-significant set-level contrast (p=0.87): the pigment and hormone
modules are not differentially deployed in dichromatic lineages in any way this
test can detect. But the test is **power-limited** — only 21 dichromatic tips,
and module selection is common across the whole panel (most tips carry at least
one selected gene), so the contingency table has little room to show
association. A negative here is "no detectable coupling," not "coupling ruled
out." The certified per-gene signals (TFAP2A, KITLG, EDN3; the steroid enzymes)
remain the substantive result; the coupling this test looks for is a stronger,
lineage-level claim the data cannot support either way.

## 5. Replay-equality assertion

Re-derives the committed result tables from the frozen inputs and asserts they
match the values recorded when the notebook was frozen — proof the replay
reproduces the checkpoint.

```{r replay-assert}
committed <- list(
  lambda = 0.6547, lambda_p = 1.10e-6,
  loss_over_gain = 9.10, dAIC_ARD_minus_ER = -19.79,
  pig_p = 0.3661, hor_p = 0.6834,
  pig_dAIC = -3.69, hor_dAIC = -5.71)
stopifnot(
  abs(lambda_res$lambda - committed$lambda) < 1e-3,
  lambda_res$p < 1e-5,
  abs(rate_res$loss_over_gain - committed$loss_over_gain) < 0.05,
  abs(rate_res$dAIC_ARD_minus_ER - committed$dAIC_ARD_minus_ER) < 0.1,
  abs(pagel_res$p[pagel_res$module == "pigmentation"] - committed$pig_p) < 1e-2,
  abs(pagel_res$p[pagel_res$module == "hormone"]      - committed$hor_p) < 1e-2)
cat("REPLAY OK — all frozen values reproduced.\n")
```
