---
title: "Module-balance selection & Opie-analog dynamics of primate hair dichromatism"
subtitle: "Frozen-data replicable notebook"
author: "Molecular Evolution analysis (pigmentation-gene-network)"
format:
  html:
    toc: true
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

## What this notebook does

Three analyses, all from **frozen inputs** in `./data/` (no network, no external
state). Any collaborator can `quarto render` this and reproduce every number and
figure.

1. **Module-balance metric** — for each independent origin of dichromatism, does
   selection tilt toward the **pigmentation** module or the **sex-hormone**
   module? Metric = `(nP − nH)/(nP + nH)` over genes under episodic selection
   (aBSREL, corrected p<0.05) along that origin's tip branches, ranging −1
   (all hormone) to +1 (all pigmentation).
2. **Per-lineage gene detail** — *which* genes in each module are selected along
   each origin, with RELAX-confirmed intensification (BH<0.05, powered origins)
   marked.
3. **Opie-analog trait dynamics** — the three tests Opie, Atkinson & Shultz
   (2012, *Commun Integr Biol* 5:458, doi:10.4161/cib.20821) ran on primate
   **mating systems** — phylogenetic signal / model choice, ancestral-state
   reconstruction, and directional transition-rate asymmetry — applied here to
   **dichromatism** using only our phenotype + tree. We hold no mating-system
   data (their supplementary values are not available), so we replicate their
   *method*, not their trait.

### Data provenance (frozen)

| file | source | contents |
|---|---|---|
| `data/branch_rates.csv` | HPC full-panel aBSREL (commit 4c07317) | 78 genes × 288 branches, per-branch ω + corrected p + selected flag |
| `data/per_origin_K.csv` | HPC per-origin RELAX (commit 4c07317) | K + BH p for 3 powered origins (Trachypithecus, Nomascus, Eulemur) |
| `data/origin_assignments.csv` | own ASR/parsimony | species → origin_id (14 genomic origins) |
| `data/gene_modules.csv` | curated panel | gene → module (pigmentation / hormone) |
| `data/species_coding.csv` | curated phenotype scoring | species → dichromatic (0/1) |
| `data/primate_species_tree.nex` | pruned primate tree | 304-tip tree for the genomic set |

Rate/ASR estimates (analysis 3) use the **full 224-tip primate phenotype tree**
(set the `PHENO_TREE_DIR` environment variable to the directory holding
`updated_pruned_tree.nex` + `primate_prelim_complete.csv`) because the
genome-sampled set is ascertainment-biased toward dichromatic species and would
inflate the gain rate. If that path
is unavailable, analysis 3 is skipped with a note.

```{r setup}
library(ape); library(phytools); library(tidyverse)
dd <- "data"
BR  <- read_csv(file.path(dd,"branch_rates.csv"), show_col_types=FALSE)
K   <- read_csv(file.path(dd,"per_origin_K.csv"), show_col_types=FALSE)
OA  <- read_csv(file.path(dd,"origin_assignments.csv"), show_col_types=FALSE)
GM  <- read_csv(file.path(dd,"gene_modules.csv"), show_col_types=FALSE)
COD <- read_csv(file.path(dd,"species_coding.csv"), show_col_types=FALSE)
gmod <- setNames(GM$module, GM$gene)
fam_order <- c("hylobatidae","cercopithecidae","cebidae","lemuridae")  # closest->furthest from human
```

## 1. Module-balance metric

```{r module-balance}
absrel_sel <- BR %>% filter(selected_flag==TRUE, is_tip==TRUE) %>%
  inner_join(OA %>% select(species, origin_id), by=c("branch"="species"))

bal <- absrel_sel %>% group_by(origin_id) %>%
  summarise(nP = n_distinct(gene[set=="pigmentation"]),
            nH = n_distinct(gene[set=="hormone"]), .groups="drop") %>%
  mutate(balance = (nP - nH)/(nP + nH), n_sel = nP + nH) %>%
  left_join(OA %>% group_by(origin_id) %>%
              summarise(family=first(family), sp=first(species),
                        n_tips=n(), .groups="drop"), by="origin_id") %>%
  mutate(family = factor(family, levels=fam_order),
         powered = if_else(origin_id %in% c("origin_7","origin_8","origin_14"),
                           "RELAX-powered","aBSREL only")) %>%
  arrange(family, balance)
knitr::kable(bal %>% select(origin_id, family, sp, nP, nH, balance, n_sel, powered),
             digits=3, caption="Module balance per origin (−1 hormone .. +1 pigmentation)")
write_csv(bal, "module_balance_results.csv")
```

The metric spans the full range: origins tilt all the way to hormone
(*Eulemur*, *Alouatta*, *Colobus guereza*) and all the way to pigmentation
(*Pithecia*). The same phenotype is reached through different parts of the
coupled network — the central finding of the study, now quantified per origin.
See `fig_module_balance.png` (built in the companion Python cell / `figures.py`).

## 2. Per-lineage gene detail

```{r per-lineage}
relax_sig <- K %>% filter(p_BH < 0.05) %>% transmute(origin_id, gene, K=round(K,2))
per_lineage <- absrel_sel %>%
  group_by(origin_id, set) %>%
  summarise(genes = paste(sort(unique(gene)), collapse=", "), .groups="drop") %>%
  pivot_wider(names_from=set, values_from=genes) %>%
  left_join(OA %>% group_by(origin_id) %>% summarise(sp=first(species), .groups="drop"), by="origin_id")
knitr::kable(per_lineage, caption="Selected genes per origin, by module (aBSREL corrected p<0.05)")
knitr::kable(relax_sig, caption="RELAX-confirmed intensification (BH<0.05, powered origins)")
```

RELAX confirms the aBSREL signal in the two powered origins with enough tips:
**origin_7 (Trachypithecus)** has 12 significant genes across both modules;
**origin_8 (Nomascus)** has POMC only; **origin_14 (Eulemur)** has none. See
`fig_per_lineage_genes.png`.

## 3. Opie-analog trait dynamics

```{r opie-analog}
pheno_dir <- Sys.getenv("PHENO_TREE_DIR", unset = "../../../../phenotype_source/data")
lk_path <- file.path(pheno_dir, "updated_pruned_tree.nex")
tr_path <- file.path(pheno_dir, "primate_prelim_complete.csv")
if (file.exists(lk_path) && file.exists(tr_path)) {
  lk_tr <- read.nexus(lk_path)
  TRAIT <- read_csv(tr_path, show_col_types=FALSE) %>%
    mutate(species=gsub(" ","_",species_binom)) %>% distinct(species,.keep_all=TRUE)
  common <- intersect(lk_tr$tip.label, TRAIT$species)
  trp <- keep.tip(lk_tr, common)
  x <- setNames(as.integer(TRAIT$hair_dichromatism_any[match(trp$tip.label, TRAIT$species)]),
                trp$tip.label)
  er  <- fitMk(trp, x, model="ER")
  ard <- fitMk(trp, x, model="ARD")
  Qm  <- as.Qmatrix(ard)
  gain <- Qm["0","1"]; loss <- Qm["1","0"]
  opie_tab <- tibble(
    opie_test = c("Model / phylogenetic signal","Gain rate (0->1)","Loss rate (1->0)",
                  "Directional asymmetry","Model support dAIC(ARD-ER)"),
    opie_2012_mating = c("lambda=0.996; RJ-MCMC best model 71% PP","strong into monogamy",
                         "zero reverse rate","INTO monogamy (absorbing)","asymmetric favored"),
    our_dichromatism = c(if_else(AIC(ard)<AIC(er),"ARD asymmetric preferred","ER"),
                         signif(gain,3), signif(loss,3),
                         paste0("OUT of dichromatism: loss ",round(loss/gain,1),"x gain"),
                         round(AIC(er)-AIC(ard),2)))
  knitr::kable(opie_tab, caption="Opie-analog: same method family, mirror-image dynamic")
  write_csv(opie_tab, "opie_analog_results.csv")
  cat(sprintf("\nFull tree: %d tips, %d dichromatic. loss:gain = %.1fx\n",
              Ntip(trp), sum(x), loss/gain))
} else {
  cat("Primate phenotype tree not found at expected path; analysis 3 skipped.\n")
}
```

**Interpretation.** Opie et al. found mating systems evolve *into* monogamy as an
absorbing state (zero exit rate). Dichromatism shows the mirror-image dynamic: a
**labile derived state**, gained rarely and lost ~9× faster (ARD preferred,
ΔAIC≈20). We reproduce their analytical approach (Mk ancestral-state
reconstruction + directional rate asymmetry) on our own trait; we do not use
their mating-system data, which is unavailable.

## 4. Sister-pair contrast (where the trait difference sits)

For each dichromatic taxon, find its **closest monochromatic relative** by
patristic distance, then list the genes under episodic selection (aBSREL) in the
dichromatic tip that its sister lacks — candidate loci for the trait difference.

```{r sister-pairs}
gtips <- unique(BR$branch[BR$is_tip]); trg <- keep.tip(tr, intersect(tr$tip.label, gtips))
dich <- COD %>% mutate(species=gsub(" ","_",species)) %>% filter(dichromatic==1) %>% pull(species)
dich_g <- intersect(trg$tip.label, dich); mono_g <- setdiff(trg$tip.label, dich)
D <- cophenetic(trg)
pairs <- lapply(dich_g, function(d){
  dd <- D[d, mono_g]; data.frame(dich=d, sister_mono=names(which.min(dd)), patristic=min(dd))
}) %>% bind_rows() %>% arrange(patristic)
gene_set_of <- function(sp) BR %>% filter(branch==sp, selected_flag==TRUE, is_tip==TRUE) %>%
  select(gene, set) %>% deframe()
contrast <- lapply(seq_len(nrow(pairs)), function(i){
  d<-pairs$dich[i]; m<-pairs$sister_mono[i]; gd<-gene_set_of(d); gm<-gene_set_of(m)
  only_d <- setdiff(names(gd), names(gm))
  data.frame(dich=d, sister_mono=m, patristic=round(pairs$patristic[i],2),
    dich_only_pig=paste(sort(only_d[gd[only_d]=="pigmentation"]),collapse=","),
    dich_only_hor=paste(sort(only_d[gd[only_d]=="hormone"]),collapse=","))
}) %>% bind_rows()
knitr::kable(contrast, caption="Genes selected in dichromatic tip but not its closest monochromatic sister")
write_csv(contrast, "sister_pair_contrast.csv"); write_csv(pairs, "sister_pairs.csv")
```

**KITLG** recurs on the pigmentation side across independent lineages (*Hylobates*,
*Trachypithecus* ×3, *Colobus*, *Erythrocebus*, *Alouatta*) — a candidate repeated
axis. *Trachypithecus francoisi* shows the richest differential (14 genes across
both modules). Two pairs (*Macaca arctoides*, *Eulemur flavifrons*) show no
differential genes — dichromatism there is not explained by selection at these
loci. Figures: `fig_sister_pair_contrast.png`, `fig_sister_network_diff.png`
(the latter maps the difference onto the STRING coupled network per pair; built
in `figures.py` / from `data/string_diff_edges.tsv`).

**Caveat:** aBSREL detects episodic selection at tips; a between-sister
difference marks lineage-specific selection, not proven causation of the phenotype.

## Session info

```{r sessioninfo}
sessionInfo()
```
